---
title: "TravToEd_queries"
author: 'Chad Kakau: 300212228'
date: "2024-12-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, include = F)
library(rJava)
library(jsonlite)
```

Use this document to retrieve census data from the Aotearoa Data Explorer site and import for analysis without downloading a file.  General process:  

1. identify key datasets
2. generate api query
3. retrieve and reshape data
4. conduct analysis
5. generate a couple of outputs.

## 1. identify key datasets  

Key datasets will be the travel to education responses from the 2018 and 2023 census years.  Try to import at a range of levels, summarised at SA1, SA2, SA3, Territorial Authority, Regional Council, and Urban/Rural.

### determine data availability  

Create a query to identify whether data is available in the database.  To request data availability, users must provide the following elements:  

  - context:
  - agencyId:
  - resourceId:
  - version:
  - key:
  - componentId:
  
Users can also provide optional information for:  

  - mode:
  - references:

```{r q_dataAvailability}
library(httr)
# url <- "https://apis.stats.govt.nz/ade-api/rest/dataflow/STATSNZ/CEN23_TRA_001/1.0?references=all&detail=full_HTTP/1.1"
url <- 'http://httpbin.org/get'
GET(url)

# first <- GET(url,
#              add_headers(accept-encoding = 'identity',
#                          Ocp-Apim-Subscription-Key = 'c5bc404c96814e9989910726559ee843')
#              )
# 
# school_data <- jsonlite::fromJSON(qry)

```


### travel to education summarised at Regional Council level 

This dataset will include data summarised at the Regional Council level and should therefor contain 17 rows.

```{r q_RC}


```

